---
title: "Converting dates, times or date-times to ISO 8601"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(sdtm.oak)
```

An SDTM DTC variable may include data that is represented in [ISO
8601](https://en.wikipedia.org/wiki/ISO_8601) format as a complete date/time, a
partial date/time, or an incomplete date/time. `{sdtm.oak}` provides the
`create_iso8601()` function that allows flexible mapping of date and time
values in various formats to a single date-time ISO 8601 format.

## Introduction

To perform conversion to the ISO 8601 format you need to pass two key arguments:

- At least one vector of dates, times, or date-times of `character` type;
- A date/time format via the `.format` parameter that instructs `create_iso8601()` on which date/time components to expect.

```{r}
create_iso8601("2000 01 05", .format = "y m d")
create_iso8601("22:35:05", .format = "H:M:S")
```

By default the `.format` parameter understands a few reserved characters:

- `"y"` for year
- `"m"` for month
- `"d"` for day
- `"H"` for hours
- `"M"` for minutes
- `"S"` for seconds

Besides character vectors of dates and times, you may also pass a single vector
of date-times, provided you adjust the format:

```{r}
create_iso8601("2000-01-05 22:35:05", .format = "y-m-d H:M:S")
```

## Multiple inputs

If you have dates and times in separate vectors then you will need to pass
a format for each vector:

```{r}
create_iso8601("2000-01-05", "22:35:05", .format = c("y-m-d", "H:M:S"))
```

In addition, like most R functions that take vectors as input,
`create_iso8601()` is vectorized:

```{r}
date <- c("2000-01-05", "2001-12-25", "1980-06-18", "1979-09-07")
time <- c("00:12:21", "22:35:05", "03:00:15", "07:09:00")
create_iso8601(date, time, .format = c("y-m-d", "H:M:S"))
```

But the number of elements in each of the inputs has to match or you will get an
error:

```{r}
date <- c("2000-01-05", "2001-12-25", "1980-06-18", "1979-09-07")
time <- "00:12:21"
try(create_iso8601(date, time, .format = c("y-m-d", "H:M:S")))
```

You can combine individual date and time components coming
in as separate inputs; here is a contrived example of year, month and day
together, hour, and minute:

```{r}
year <- c("99", "84", "00", "80", "79", "1944", "1953")
month_and_day <- c("jan 1", "apr 04", "mar 06", "jun 18", "sep 07", "sep 13", "sep 14")
hour <- c("12", "13", "05", "23", "16", "16", "19")
min <- c("0", "60", "59", "42", "44", "10", "13")
create_iso8601(year, month_and_day, hour, min, .format = c("y", "m d", "H", "M"))
```

The `.format` argument must be always named; otherwise, it will be treated as if
it were one of the inputs and interpreted as missing.

```{r}
try(create_iso8601("2000-01-05", "y-m-d"))
```


## Format variations

The `.format` parameter can easily accommodate variations in the format of the
inputs:

```{r}
create_iso8601("2000-01-05", .format = "y-m-d")
create_iso8601("2000 01 05", .format = "y m d")
create_iso8601("2000/01/05", .format = "y/m/d")
```

Individual components may come in a different order, so adjust the format
accordingly:

```{r}
create_iso8601("2000 01 05", .format = "y m d")
create_iso8601("05 01 2000", .format = "d m y")
create_iso8601("01 05, 2000", .format = "m d, y")
```

All other individual characters given in the format are taken strictly, e.g.
the number of spaces matters:

```{r}
date <- c("2000 01 05", "2000  01 05", "2000 01  05", "2000   01   05")
create_iso8601(date, .format = "y m d")
create_iso8601(date, .format = "y  m d")
create_iso8601(date, .format = "y m  d")
create_iso8601(date, .format = "y   m   d")
```

The format can include regular expressions though:

```{r}
create_iso8601(date, .format = "y\\s+m\\s+d")
```

By default, a streak of the reserved characters is treated as if only one was
provided, so these formats are equivalent:

```{r}
date <- c("2000-01-05", "2001-12-25", "1980-06-18", "1979-09-07")
time <- c("00:12:21", "22:35:05", "03:00:15", "07:09:00")
create_iso8601(date, time, .format = c("y-m-d", "H:M:S"))
create_iso8601(date, time, .format = c("yyyy-mm-dd", "HH:MM:SS"))
create_iso8601(date, time, .format = c("yyyyyyyy-m-dddddd", "H:MMMMM:SSSS"))
```

### Multiple alternative formats

When an input vector contains values with varying formats, a single format may
not be adequate to encompass all variations. In such situations, it's advisable
to list multiple alternative formats. This approach ensures that each format is
tried sequentially until one matches the data in the vector.

```{r}
date <- c("2000/01/01", "2000-01-02", "2000 01 03", "2000/01/04")
create_iso8601(date, .format = "y-m-d")
create_iso8601(date, .format = "y m d")
create_iso8601(date, .format = "y/m/d")
create_iso8601(date, .format = list(c("y-m-d", "y m d", "y/m/d")))
```

Consider the order in which you supply the formats, as it can be significant. If
multiple formats could potentially match, the sequence determines which format
is applied first.

```{r}
create_iso8601("07 04 2000", .format = list(c("d m y", "m d y")))
create_iso8601("07 04 2000", .format = list(c("m d y", "d m y")))
```

## Parsing of date or time components

By default, date or time components are parsed as follows:

- year: either parsed from a two- or four-digit year;
- month: either as a numeric month (single or two-digit number) or as an English abbreviated month name (e.g. Jan, Jun or Dec) regardless of case;
- month day: are parsed from two-digit numbers;
- hour and minute: are parsed from single or two-digit numbers;
- second: is parsed from single or two-digit numbers with an optional fractional part.

```{r}
# Years: two-digit or four-digit numbers.
years <- c("0", "1", "00", "01", "15", "30", "50", "68", "69", "80", "99")
create_iso8601(years, .format = "y")

# Adjust the point where two-digits years are mapped to 2000's or 1900's.
create_iso8601(years, .format = "y", .cutoff_2000 = 20L)

# Both numeric months (two-digit only) and abbreviated months work out of the box
months <- c("0", "00", "1", "01", "Jan", "jan")
create_iso8601(months, .format = "m")

# Month days: single or two-digit numbers, anything else results in NA.
create_iso8601(c("1", "01", "001", "10", "20", "31"), .format = "d")

# Hours
create_iso8601(c("1", "01", "001", "10", "20", "31"), .format = "H")

# Minutes
create_iso8601(c("1", "01", "001", "10", "20", "60"), .format = "M")

# Seconds
create_iso8601(c("1", "01", "23.04", "001", "10", "20", "60"), .format = "S")
```
