---
title: "All about Metadata"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{All about Metadata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{css, echo = FALSE}
#pkgdown-sidebar{
 display:none;
}
.wide .table{
  font-size: 8px;
  overflow: visible
}

```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

{sdtm.oak} is a metadata-driven solution that is designed to be Electronic Data 
Capture (EDC) and standards agnostic. Throughout this article, the term "metadata" 
is used several times. In this context, "metadata" refers to the specific 
metadata used by {sdtm.oak}. This article aims to provide users with a more 
detailed understanding of the {sdtm.oak} metadata.

In general, metadata can be defined as "data about data." It does not include 
any patient-level data. Instead, the metadata provides a blueprint of the data 
that needs to be collected during a study.

## Standards Metadata

The standards metadata used in {sdtm.oak} is sourced from the CDISC Library or 
sponsor MDR or any other form of documentation where standards are maintained. 
This metadata provides information on the following:

- The relationship between Data Collection Standards (eCRF & eDT), SDTM mapping, 
and Controlled Terminology
- Machine-readable standard SDTM mappings
- Algorithms and associated metadata required for the SDTM automation of 
standards in the study.

In the upcoming releases of {sdtm.oak}, we will effectively utilize the 
standards metadata and customize it to meet the study requirements.

## Study Definition Metadata

Study Definition Metadata is also referred to as Study Metadata. Study Definition 
Metadata provides information about the eCRF and eDT data collected in the study.

**eCRF Metadata**
The eCRF Design Metadata is fetched from the EDC system. This Metadata includes

- Forms Metadata: Identifier, eCRF label, Repeating format and other properties of the eCRF.

- Fields Metadata: Identifier, question label, datatype, and other 
properties of data collection fields in the study.

- Data Dictionaries: Identifier and the controlled terms collected at the
source.

- Visits: Name of the visits as defined in the EDC.

**eDT Metadata**

eDT Metadata is the blueprint metadata that describes the data collected as part of that external data transfer. This includes

- Dataset name, label, repeating properties, etc.

- Variable name, datatype, label and associated codelist, etc.


## Study SDTM Mappings Metadata (specifications)

Study SDTM mappings metadata is the study SDTM specification. To develop the SDTM domains, {sdtm.oak} requires the user to prepare the Study SDTM mappings metadata. Unlike the conventional SDTM specification, which includes one tab per domain defining the target (SDTM domain, Variables) to source (raw dataset, raw variables) and SDTM mappings, the SDTM spec for {sdtm.oak} defines the source-to-target relationship. For each source, the SDTM mapping, algorithms, and associated metadata are defined. The table below presents the columns in the SDTM mapping specification and its explanation.

<div class="wide">
```{r echo = FALSE, results = "asis"}
library(knitr)
definition <- data.frame(
  Variable_Name = c(
    "study_number",
    "raw_source_model",
    "raw_dataset",
    "raw_dataset_ordinal",
    "raw_dataset_label",
    "raw_variable",
    "raw_variable_label",
    "raw_variable_ordinal",
    "raw_variable_type",
    "raw_data_format",
    "raw_codelist",
    "study_specific",
    "annotation_ordinal",
    "mapping_is_dataset",
    "annotation_text",
    "target_domain",
    "target_sdtm_variable",
    "target_sdtm_variable_role",
    "target_sdtm_variable_codelist_code",
    paste("target_sdtm_variable_",
    "controlled_terms_or_format"),
    "target_sdtm_variable_ordinal",
    "origin",
    "mapping_algorithm",
    "sub_algorithm",
    "target_hardcoded_value",
    "target_term_value",
    "target_term_code",
    "condition_ordinal",
    "condition_group_ordinal",
    "condition_left_raw_dataset",
    "condition_left_raw_variable",
    "condition_left_sdtm_domain",
    "condition_left_sdtm_variable",
    "condition_operator",
    "condition_right_text_value",
    "condition_right_sdtm_domain",
    "condition_right_sdtm_variable",
    "condition_right_raw_dataset",
    "condition_right_raw_variable",
    "condition_next_logical_operator",
    "merge_type",
    "merge_left",
    "merge_right",
    "merge_condition",
    "unduplicate_keys",
    "groupby_keys",
    "target_resource_dataset",
    "target_resource_variable"
  ),
  `Description_of_the_variable` = c(
    "Study Number",
    "Data Collection model",
    "Name of the raw or source dataset",
    "Ordinal of the raw dataset as defined in EDC or eDT specification",
    "Label of the raw or source dataset",
    "Name of the raw variable",
    "Label of the raw variable",
    paste(
      "Ordinal of the variable as defined in the eCRF or",
      "eDT specification"
    ),
    "Type of the Raw Variable",
    "Data format of the raw variable",
    paste(
      "Dictionary name which is assigned to the ",
      "eCRF field or a eDT variable"
    ),
    paste(
      "`TRUE` indicates that the source is study specific. ",
      "`FALSE` indicates that the raw variable is part of data standards"
    ),
    "Ordinal of the SDTM mappings for the particular raw source",
    paste(
      "Indicates if the SDTM mapping is at the dataset level. ",
      "`TRUE` indicates that it is dataset level mapping."
    ),
    "SDTM mapping text or annotation text",
    "Name of the target domain.",
    "Name of the target SDTM variable",
    "CDISC Role for the SDTM target variable defined in the annotation.",
    paste(
      "NCI or sponsor code of the codelist assigned to the ",
      "SDTM target variable defined in the annotation."
    ),
    paste(
      "Controlled terms or format for the target variable ",
      "defined in the annotation (as defined per CDISC).",
      "`target_sdtm_variable_controlled_terms_or_format` is required ",
      "for SDTM Define.xml"
    ),
    "Ordinal of the target SDTM variable",
    "Origin of metadata source, values are subject to controlled terminology",
    "Mapping Algorithm",
    "The sub-algorithm (scenario) of the source-to-target mapping",
    "Text (Hardcoded value) that applies to the target.",
    paste(
      "CDISC Submission value or sponsor value which represents a",
      "hardcoded text"
    ),
    paste(
      "NCI code or sponsor code of the hardcoded value"
    ),
    paste(
      "Ordinal of a (sub)condition, increasing when there ",
      "are more than one sub-conditions (e.g. X AND Y)"
    ),
    paste(
      "Ordinal of a group of sub-conditions, used to ",
      "disambiguate complex conditions such as (A AND B) OR C. ",
      "The ordinal increases in each group and gives the final ",
      "precedence of the logical operators."
    ),
    "Name of the raw dataset on the left part of the condition",
    "Name of the raw variable on the left part of the condition",
    "Name of the SDTM variable used in the left part of the condition.",
    paste(
      "Name of the SDTM domain of the variable that is used in ",
      "the left part of the condition."
    ),
    "Operator between the left and right part of the condition",
    paste(
      "A text that applies to the right part of the condition as ",
      "indicated per `condition_operator`."
    ),
    "Name of the SDTM variable used in the right part of the condition.",
    paste(
      "Name of the SDTM domain of the variable that is used ",
      "in the right part of the condition."
    ),
    "Name of the raw dataset on the right part of the condition",
    "Name of the raw variable on the right part of the condition",
    paste(
      "The logical operator that applies to the next ",
      "sub-conditions, typically AND, OR"
    ),
    "Specifies the type of join",
    "Specifies the left component of the merge",
    "Specifies the right component of the merge",
    paste(
      "Specify the condition of the join (e.g. a specific ",
      "variable that should match in the components of the merge)"
    ),
    paste(
      "Raw variables that should be used to determine whether ",
      "an observation in the source data is a duplicate record and ",
      "subject to being removed"
    ),
    paste(
      "Raw Variables or aggregation functions (i.e. earliest, ",
      "latest) to group source data records before mapping to SDTM"
    ),
    paste(
      "Raw dataset name of the raw variable. This will be used when ",
      " values are assigned from a from a different source",
      "other than the source the mapping is associated with"
    ),
    paste(
      "Raw variable name. This will be used when ",
      "values are assigned from a from a different source",
      "other than the source the mapping is associated with"
    )
  ),
  Example_Values = c(
    "test_study",
    "e-CRF or eDT",
    "VTLS1, DEM",
    "1, 2, 3, etc",
    "Vital Signs,<br> Demographics",
    "SEX_001, <br> BRTHDD",
    "Systolic Blood Pressure,<br>Birth Day",
    "1, 2, 3, etc",
    "Text Box,<br> Date control",
    "$200,<br> dd MON YYYY",
    "SEX, ETHNIC",
    "TRUE, FALSE",
    "1, 2, 3, etc",
    "TRUE, FALSE",
    "VS.VSORRES when VSTESTCD = 'SYSBP'",
    "VS, MH",
    "VSORRES, MHSTDTC",
    "Topic Variable,<br>Grouping Qualifier,<br>Identifier Variable",
    "C66742<br>C66790",
    "(AGEU)<br>ISO 8601<br>(SEX)",
    "1, 2, 3",
    "Derived, <br>Assigned, <br>Collected, <br>Predecessor",
    "DATASET_LEVEL<br>ASSIGN_CT<br>AE_AEREL<br>HARDCODE_CT",
    "ASSIGN_NO_CT<br>HARCODE_CT",
    "ALZHEIMER'S DISEASE HISTORY",
    "Y, <br>beats/min, <br>INFORMED CONSENT OBTAINED",
    "C49488",
    "1, 2, 3",
    "1, 2, 3",
    "VTLS1",
    "POSITION",
    "AE",
    "AEENRTPT",
    paste(
      "Checked<br>Not_checked<br>Is_null<br>Is_not_null",
      "<br>Equals_to<br>Different_to<br>is_numeric<br>in",
      "<br>not_in"
    ),
    "('Not Recovered/Not Resolved','Recovering/Resolving')<br>HOSPITALIZATION",
    "AE",
    "AETERM",
    "SMKHX",
    "SUNAM",
    "and, or",
    "left_join<br>right_join<br>full_join<br>visit_join<br>subject_join",
    "VTLS1",
    "VACREC",
    "VTLS1.SUBJECT = VACREC.SUBJECT,<br>MD1.MDNUM = VACREC.MDNUM",
    "VTLS1.SUBJECT,<br>VTLS1.DATAPAGEID",
    "TXINF1.DATAPGID, <br>Earliest",
    "AEDE",
    "DATAPAGEID"
  ),
  Association_with_mapping_Algorithms = c(
    "Generic Use",
    "Generic Use",
    "Required for all mapping algorithms",
    "Generic Use",
    "Generic Use",
    "Generic Use",
    "Generic Use",
    "Generic Use",
    "Required for all mapping algorithms",
    "Required for all mapping algorithms",
    "Required for all mapping algorithms",
    "Generic Use",
    "Required for all mapping algorithms",
    "Required for all mapping algorithms",
    "Generic Use",
    "Required for all mapping algorithms",
    "Required for all mapping algorithms",
    "Required for all mapping algorithms",
    "Required for all mapping algorithms",
    "Generic Use",
    "Required for all mapping algorithms",
    "Used for define.xml",
    "Required for all mapping algorithms",
    "Only when Mapping Algorithm is <br>IF_THEN_ELSE<br>DATASET_LEVEL",
    "ASSIGN_NO_CT<br>HARDCODE_NO_CT",
    "HARDCODE_CT",
    "IF_THEN_ELSE",
    "IF_THEN_ELSE",
    "IF_THEN_ELSE",
    "IF_THEN_ELSE",
    "IF_THEN_ELSE",
    "IF_THEN_ELSE",
    "IF_THEN_ELSE",
    "IF_THEN_ELSE",
    "IF_THEN_ELSE",
    "IF_THEN_ELSE",
    "IF_THEN_ELSE",
    "IF_THEN_ELSE",
    "IF_THEN_ELSE",
    "IF_THEN_ELSE",
    "MERGE",
    "MERGE",
    "MERGE",
    "MERGE",
    "REMOVE_DUP",
    "GROUP_BY",
    "ASSIGN_NO_CT",
    "ASSIGN_NO_CT"
  ),
  stringsAsFactors = TRUE
)
knitr::kable(definition, col.names = gsub("_", " ", names(definition)))
```
